---
title: "Data Visualization - Mini-Project 1"
author: "Omar Abdallah Mateo `oabdallahmateo8279@floridapoly.edu`"
output:
  html_document:
    df_print: paged
---

## Load Data

```{r load-clean-data, warning=FALSE, message=FALSE}
library(tidyverse)
us_births <- read_csv("https://raw.githubusercontent.com/reisanar/datasets/master/us_births_00_14.csv", col_types = cols())

```

## Summary

```{r plot-summary}
library(dplyr)

births_summary <- us_births %>%
  group_by(year) %>%
  summarize(total_births = sum(births))

# View the summarized data
print(births_summary)

```

## Visualizations

```{r line chart}
library(ggplot2)

# Line chart
ggplot(births_summary, aes(x = year, y = total_births)) +
  geom_line(color = "steelblue") +
  labs(x = "Year", y = "Total Births", title = "Births by Year - Line Chart")

```

```{r bar chart}
library(ggplot2)
library(scales)
# Convert month to a factor with all levels
us_births$month <- factor(us_births$month, levels = unique(us_births$month),
                          labels = c("January", "February", "March", "April", "May", "June",
                                     "July", "August", "September", "October", "November", "December"))

# Calculate total births by month
births_summary <- us_births %>%
  group_by(month) %>%
  summarize(total_births = sum(births))

# Reorder the levels of the month factor based on total births in descending order
births_summary <- births_summary[order(births_summary$total_births, decreasing = TRUE), ]
us_births$month <- factor(us_births$month, levels = births_summary$month)

# Bar chart of births by month
ggplot(us_births, aes(x = month, y = births)) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(labels = label_number(suffix = "")) +
  labs(x = "Month", y = "Number of Births", title = "Births by Month") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


```

```{r boxplot}
library(ggplot2)

# Convert month to a factor with all levels
us_births$month <- factor(us_births$month, levels = unique(us_births$month),
                          labels = c("January", "February", "March", "April", "May", "June",
                                     "July", "August", "September", "October", "November", "December"))

# Create a boxplot of births by month
ggplot(us_births, aes(x = month, y = births)) +
  geom_boxplot(fill = "steelblue") +
  labs(x = "Month", y = "Number of Births", title = "Distribution of Births by Month") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

```{r heatmap}
library(reshape2)
library(dplyr)

# Convert month and day_of_week to factors with all levels
us_births$month <- factor(us_births$month, levels = unique(us_births$month),
                          labels = c("January", "February", "March", "April", "May", "June",
                                     "July", "August", "September", "October", "November", "December"))
us_births$day_of_week <- factor(us_births$day_of_week, levels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

# Summarize births by month and day_of_week
births_summary <- us_births %>%
  group_by(month, day_of_week) %>%
  summarize(total_births = sum(births))

# Reshape the data for creating the heatmap
births_matrix <- dcast(births_summary, day_of_week ~ month, value.var = "total_births")

# Create the heatmap
ggplot(data = births_summary, aes(x = month, y = day_of_week, fill = total_births)) +
  geom_tile() +
  labs(x = "Month", y = "Day of the Week", title = "Number of Births by Month and Day of the Week") +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "right")


```

# Report

## What were the original charts you planned to create for this assignment?

These were the original charts; however, I was facing some issues with the bar chart with the labels, but it is good now.

## What story could you tell with your plots?

I found out that the highest peak of births was in 2007; however, after that, it decreased drastically. It will be interesting to compare this with the family's economy since I feel it might be related to the economic crisis of 2008.

I also searched throw the months to find out what is the month with the highest number of births. I have always heard that October/November has the highest number of births because of St. Valentine's Day in February; after looking at the graph, I can confirm that this legend is false. The month in which most births occur is August.

## How did you apply the principles of data visualization and design for this assignment?

I tried to use the same color for each visualization to keep the same theme. And I used horizontal bar charts instead of vertical ones and organized them from bigger to smaller because it is way easier to read them.
